The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-08-07 paper.qmd
finished the testing signal suppression section

## What can and cannot be suppressed?
The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. Specifically that uncertainty and the purpose of visualisation are somewhat at odds with one another. There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation using it to extract several signals. If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. Unfortunately it is not clear whether or not uncertainty visualisations for EDA are possible.

@@

Some authors recognize inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020]. While we agree that people cannot prevent themselves from performing inference, this does not mean it is possible to include uncertainty in a visualisation designed for EDA. However, this does mean we should endeavor for a versatile uncertainty visualisation method that is able to perform signal-suppression on all the signals displayed in the visualisation. It is unlikely that the methods discussed thus far are able to achieve this.

@@

All of the individual t-tests of fail to reject the null hypothesis that the coefficients are different from 0. We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. We then do a comparison of two means t-test and find that several of the values need to be visually distinguishable from each other. The VSUP method must pick a single color for each estimate, and these colours must be *either* visually distinguishable or indistinguishable from each other. We cannot perform signal-suppression on both these signals simultaneously.

@@

### Value extraction of uncertainty statistics
Uncertainty visualisations are most commonly evaluated on how accurately readers can extract information from them [@Hullman2019]. This most commonly comes in the form of separate questions about estimates variance [@Kinkeldey2014]. This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". Additionally, uncertainty visualisation experiments often compare graphics on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020] regardless of what information can actually be extracted by the chart. This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? Uncertainty is not something we are trying to infer, uncertainty is a by-product of inference. By shifting the focus of our inference to $Var(X)$ or $P(X)$ we are just evaluating visualisations on their ability to convey uncertainty statistics. These graphics may not always set out to communicate uncertainty as a signal, but they are certainty evaluating uncertainty as a signal.

@@

Directly asking about uncertainty evaluates uncertainty as a signal, not as noise, so it is not a reliable evaluation method when we are designing graphics for signal suppression. Many uncertainty visualisation researchers are already aware of this, so direct questions are not the only methods used to evaluate uncertainty visualisations. We will discuss these alternatives in the following section.

@@

Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy". This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions.

The disconnect between trust and transparency extends to evaluation experiments as well. @Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust. They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive. The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy. However, this perspective conflates transparency and trustworthiness. If a model has very high uncertainty because of a very low sample size, the heuristics a human might use to make the same decision could be more accurate.

Unfortunately, experiments where the best response comes from participants who outright ignore the uncertainty information are not uncommon. A simple example of this is experiments where participants are expectd to extract the mean of the distribution. In these cases, we would expect the best performing visualisation to be one that does not include uncertainty at all, as it would simply show the mean. Similarly, many uncertainty visualisations are evaluated on how much risk aversion the participants display [@Hullman2019]. Additionally, comparing participant responses to that of a ration agent has been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational]. This standard seems to go against the purpose of uncertainty visualisation as a rational agent, *by definition* should *ignore* uncertainty information. Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information. The only cases where a rational agent should not ignore the uncertainty information is when it is required to optimise their utility function, which would mean the uncertainty is not acting as noise, it is acting as signal.

@@

For these reasons we do not believe trust, confidence, or risk-aversion are useful measures to evaluate uncertainty visualisations. While they are designed to capture the secondary effects of uncertainty, using them as primary measures of visualisation performance incentives the use of visualisations that *do not* include uncertainty.

@@

While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results. Additionally, by having a ground truth, they end up implicitly asking users to either treat uncertainty as a signal, or to ignore it entirely, neither of which is advisable for evaluating graphics on signal-suppression. So, if we cannot ask direct questions about uncertainty, *and* we can't use uncertainty statistics as ground truths in our experiments, how *should* we evaluate uncertainty?

## Testing signal supression

If a graphic was performing signal suppression we would expect the values to be harder to read at higher levels of uncertainty, something that is often seen as a bad thing in current evaluation studies.
- trust experiment
- weird animation of grass

Since the primary goal of visualisation *is* developing insights [@North2006], you will probably extract several signals and use them to make a broader conclusion. A simple example is the spatial trend in

If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?
 When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

So, the current methods of measuring or understanding the role of uncertainty in a visualisation is questionable at best, however this is not because visualisation authors are missing the mark, but rather uncertainty is *particularly* difficult to express in a visualisation. In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying. Unlike statistical models, visualisations are used in both data exploration and communication. This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is. Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation. Two people looking at the same visualisation might, just by chance, develop two entirely different insights and draw inference on two completely different statistics. These unique and fascinating challenges that are faced by uncertainty visualisation have been completely untouched by the literature. This section will cover some interesting research in uncertainty visualisation and suggestions for better ways to measure uncertainty.

### Qualitative Studies
Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses. This method can enlighten authors as to *how* the uncertainty information was used by the participants. @Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper. @Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making. The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty. This study was an excellent example in a useful experimental design. They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations. Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations. This experimental framework could replicate this process.

### Just noticeable signal

there are some papers that do identify signal suppress effects, but they discuss it as a failure rather than a success.

It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered. The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.

@Patrick2023 found that human viewers were less sensitive to deviations from the null hypothesis than the typical statistical tests.

What this all means, is that there will always be some trade off when we are designing a visualisation for signal-suppression. There will be a trade off in *which* signal we suppress, as well as *how much* we suppress it. This is true for any graphic that is comparable in some way to a statistical test or process. A good example is the line-up protocol, which is a visualisation tool that used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. This motivation is similar to our signal-suppression goal. @Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing. In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference. While these methods apply different statistical philosophies, they both investigated the sensitivity of visualisation methods relative to existing evaluation methods. This approach may prove useful when considering

## Identifying heuristics
- Things that are true for both normal visualisation and uncertainty, but of particular interest for uncertainty

Additionally, these heuristics and biases can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].

Heuristic checks are useful because they look at unknown pitfalls that might exist in interpretation of current plots. Since the hypothesis for these experiments are usually quite specific, e.g. "do people perceive a an outcome that is within the bar as more likely than one outside it, even if both outcomes are the same distance from the mean?". This means they are less likely to fall into the trap of trying to answer questions that are *far* too broad to be answered with a single experiment (e.g. "is a scatter plot better at showing uncertainty than a box plot"). This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016]. @Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars. Similar effects have been identified in other types of uncertainty displayed. @Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution. The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].

In a similar vein, experiments that verify smaller aspects of plot design might be more useful to the field in the long run because it helps contribute to a larger working theory of "how do we see visualisations". Many visualisation experiments try to compare two plots with several differences, but do not seem to be interested in the mechanisms by which we extract information from visualisations. Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.

>>>>

## What can and cannot be suppressed?
The methods used by the VSUP bring to light a slight problem with uncertainty visualisation. Specifically that uncertainty and the purpose of visualisation are somewhat at odds with one another. There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation using it to extract several signals. If we are designing an uncertainty visualisation for communication then we can just suppress the specific signal we are seeking to communicate. In the map example, we would consider @fig-vsup to be a success as the only signal we are concerned with is the spatial trend. However, it is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006]. For uncertainty visualisation for EDA to work, we would need to assume that suppressing individual estimates using their variance should naturally extend to broader suppression of plot level insights. Unfortunately it is not clear whether or not uncertainty visualisations for EDA are possible.

@@

Some authors recognize inference will occur (in some shape or form) and believe uncertainty *should* be visualised but do not recognize *how* uncertainty would be visualised. @Hullman2021 argued that there is no such thing as a "model-free" visualisation, therefore all visualisations require uncertainty as we are always performing inference. However, even something as simple as calculating the uncertainty that is used to suppress our visualisation is not model free, as we need to identify if the sampling variance or the sample variance is more appropriate [@Hofman2020]. While we agree that people cannot prevent themselves from performing inference, this does not mean it is possible to include uncertainty in a visualisation designed for EDA. However, this does mean we should endeavor for a versatile uncertainty visualisation method that is able to perform signal-suppression on all the signals displayed in the visualisation.

Uncertainty visualisation for EDA would be possible if we designed a plot in such a way that suppressing individual estimates using their variance would naturally extend to broader suppression of plot level insights. This assumption is commonly made by visualisation researchers in normal visualisation experiments [@North2006], however achieving it would largely depend on the methods use to perform signal suppression. By manually combining the values in @fig-vsup, we violated this requirement, so the VSUP is not versatile enough to act as an uncertainty visualisation for EDA.

@@

All of the individual t-tests of fail to reject the null hypothesis that the coefficients are different from 0. We then make a visualisation that suppresses this signal and ensures that all of the estimates are visually indistinguishable from 0. We then do a comparison of two means t-test and find that several of the values need to be visually distinguishable from each other. The VSUP method must pick a single color for each estimate, and these colors must be *either* visually distinguishable or indistinguishable from each other. We cannot perform signal-suppression on both these signals simultaneously.

@@

### Value extraction of uncertainty statistics
Uncertainty visualisations are often evaluated based on how accurately [@Hullman2019] viewers can separately extract the estimate and the variance [@Kinkeldey2014]. This means a significant chunk of evaluation studies boil down to showing a participant a visualisation and asking questions such as "what is the variance of $X$?", or "what is the mean of $X$?". This seems like a relatively straight forward approach, and it is similar to how non-uncertainty visualisations are evaluated, but is this appropriate for uncertainty visualisations? The role of uncertainty is rarely evaluated in these studies as the graphics are often compared on the basis of being "uncertainty visualisations" [@Ibrekk1987; @Hullman2015; @Hofman2020], a class that has no established definition. By shifting the focus of our inference from $\bar{X}$ to $Var(X)$ or $P(X)$ we end up evaluating visualisations on their ability to convey uncertainty statistics, rather than on uncertainties ability to suppress statistics. This leads to a series of experiments where the uncertainty is evaluated as a signal even if that was not the goal of the experiment.

@@

The biggest failing of this evaluation method is not the predictable outcomes, but that it encourages us to see successful examples of signal-suppression as a failing. @Blenkinsop2000 commented that visually integrable depictions of uncertainty should be avoided, as they decrease the viewers confidence in their extracted data values. This conclusion is antithetical to the goals of signal-suppression and occurs because these methods evaluate uncertainty as a signal, not as noise.

@@

Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication. Evaluating visualisations on trust conflates trust and transparency and ultimately discourages signal-suppression. We can see this effect pop up in the results of evaluation studies. For example, @Zhao2023 found that participants were more trusting of model estimates with low uncertainty, but this effect did not carry over to estimates with high uncertainty. Despite decreased trust being a desired outcome of signal suppression, the authors discussion implied this result was not desired. This perspective ends up extending to visualisation authors as well. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy". This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions. This is the end result of designing uncertainty visualisations for increased trust, rather than signal-suppression.

@@

Risk-aversion is another secondary effect of uncertainty visualization that is used to evaluate uncertainty visualisations [@Hullman2019]. Risk aversion is an economics term used to describe an agent who would chose a random variable with a lower expected payout because it also has a lower variance. Risk aversion is considered irrational behaviour because it is a deviation from the behavior of a rational agent. Comparing participant responses to that of a ration agent has even been suggested as a benchmark for uncertainty visualization experiments [@wu2023rational]. However, just like trust, evaluating graphics on risk-aversion discourages signal-suppression. Rational agents *by definition* should *ignore* uncertainty information. Risk aversion is considered to be irrational because it means the economic agent *is* considering the uncertainty information. The only cases where a rational agent should not ignore the uncertainty information is when the uncertainty is signal, not noise. Designing graphics that encourage choices that align with that of a rational agent, is to encourage graphics that do not include uncertainty at all.

For these reasons we do not believe trust, confidence, or risk-aversion are useful measures to evaluate uncertainty visualisations. While they are designed to capture the secondary effects of uncertainty, using them as primary measures of visualisation performance is in direct conflict with designing visualisations for signal-suppression.

@@

While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results. Additionally, by having a ground truth, they end up implicitly asking users to either treat uncertainty as a signal, or to ignore it entirely, neither of which is advisable for evaluating graphics on signal-suppression.

## Testing signal supression
If we cannot ask direct questions about uncertainty, we cant measure the secondary effects of uncertainty, and we can't ask indirect questions about the uncertainty, how are we supposed to evaluate uncertainty visualisations? How do you measure something that disappears the second you look directly at it? To evaluate uncertainty visualisations, we need an experimental design that evaluates uncertainty as noise, not as signal. This means we need to measure *uncertainty's impact on the signal*, not the uncertainty itself.

### Comparing to hypothesis tests
The most obvious way to evaluate uncertainty visualizations is to compare the visualisations to statistical tests. If a graphic was performing signal-suppression we would expect the signal to be harder to see at higher levels of uncertainty. The ideal outcome is a an uncertainty visualisation where the signal is only perceivable when it would be identified by a hypothesis test. Evaluating visualisations as though they are akin to hypothesis tests is a well established concept in visualisation. The lineup protocol is a good example of this approach, which is a confirmatory visualisation tool that can be used to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010]. The motivation behind the lineup protocol is similar to the motivations behind signal-suppression, although the lineup protocol is more explicitly tied to a specific hypothesis. @Patrick2023 compared standard statistical tests to the lineup-protocol, and evaluated the visualisations using the power curves that are typical for hypothesis testing. In a similar vein, @Kim2019 investigated how different uncertainty visualisation methods influenced user's prior beliefs, and evaluated the graphics by comparing their results to those from Bayesian inference. This approach is similar to the lineup protocol as it also evaluates a visualisation by comparing it to an analogous statistical calculation, however it does so using a different statistical philosophy.

We can use a similar evaluation method to evaluate uncertainty visualisations, however, special care would need to be taken for their versatility requirement. While lineup protocols often have a specific hypothesis that is used to create the null distribution, uncertainty visualisations are intended to be more flexible and can be created without visualizing a specific null distribution. Human viewers using the lineup protocol are less sensitive to deviations from the null hypothesis than the typical statistical tests [@Patrick2023]. This means swapping to a visual format and giving viewers the ability to check several hypothesis at once can reduce the effectiveness of the hypothesis test. Therefore it is highly likely that uncertainty visualisations would not produce identical results to standard hypothesis tests or even the lineup-protocol. Additionally, uncertainty visualisations would need to be simultaneously checked against several hypothesis that come from a variety of null distributions.

### Qualitative Studies
Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses. This method can enlighten authors as to *how* the uncertainty information was used by the participants. @Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", even though this qualitative analysis did not make it into the final paper. @Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making. The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty. They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and identified where in the decision making process participants made these considerations.

### Heuristics
While experiments that explicitly identify heuristics in current methods are not technically measuring signal-suppression, they are still a useful consideration to keep in mind when designing experiments for signal suppression. Heuristic checks are look at unknown pitfalls that might exist in interpretation of current plots [@Hullman2016] and can change depending on the larger scope of the graphic and the population we are communicating with [@Spiegelhalter2017; @Kinkeldey2014].

Several heuristics are of particular importance for uncertainty visualisations, and are likely to impact how well different methods performing signal suppression. The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]. Points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on an outcome was closer to the center of the distribution (and therefore more likely) [@Padilla2017]. This can be considered an extension of the within bar bias, where participants looking at a bar chat with error bars view outcomes within the bar as more likely than those outside it [@Newman2012]. Several studies have found that viewers use a heuristic where they compare the distance between two estimates to estimate if they are different, and in doing so, ignore the uncertainty information [@uncertchap2022].

These heuristics have the potential to create noise in signal suppression evaluations, and should be kept in mind when designing an evaluation experiment.
